Distance Sampling in \proglang{R}

### 2016-02-01 paper.Rmd
A few more edits... now I'm finished
I have left a few comments in, they appear in italics in the pdf and
inside _(LM … )_ in the .Rmd so are pink. I have commented out Len’s
comments as I have tried to address them but they are still in the .Rmd
file in <!— —> and appear in blue. Actually colours may vary depending
on your personal setup.

Until now those wishing to use our \proglang{R} packages for straight forward distance sampling analyses would have had to negotiate the complex package \pkg{mrds} [@mrds-pkg] designed for mark-recapture distance sampling. In addition, \pkg{mrds} requires a complex data structure before analyses can be completed. \pkg{Distance} is a wrapper package around \pkg{mrds} making it easier to get started with basic distance sampling analyses in \proglang{R}. Like the histogram function, the most basic detection function estimation only requires a numeric vector of distances. Here we demonstrate how to use \pkg{Distance} to fit detection functions, perform model checking and selection, and estimate abundance and/or density.

@@

Given that we can assume these particular availability function there are a number of ways we can use either the estimated detection function or the composite function we fit to our data to estimate abundance. One simple idea begins with finding the area under the composite function we fit to our data out to a truncation distance $w$. If we then find the area under the availability function out to distance $w$, then dividing the former area by the latter gives us an estimate of the propoertion of objects seen within distance $w$ of the transect. We can then scale up the number of objects we saw, first to give us an estimate of how many objects were within distance $w$ of the transects and then again by the proportion of the study area we have sampled to give an estimate of abundance for the entire study area.

```{r detectdists, fig.height=5.5, fig.width=8, echo=FALSE, message=FALSE, fig.cap="Top panels show an example detection function (left), the expected avaiability of objects (middle) and the resulting destribution of detections (right) for line transects. Bottom panels show an example detection function (left), the expected avaiability of objects (middle) and the resulting destribution of detections (right) for point transects. The distributions in the right hand plots are obtained by multiplying the detection function by the availability function and are referred to in the text as the 'composite function'. \\label{fig:detectiondist}"}

@@

Figure \ref{fig:pointslines} shows simulated sampling of a population of 500 objects using line and point transects and their corresponding histograms of observed detection distances. Note that for the purposes of distance sampling an object may either refer to an individual in a population or a cluster of individuals. In the latter instance, the density or abundance of clusters is estimated separately to the mean cluster size which can later be used to estimate the density or abundance of individuals.

@@

\pkg{Distance} provides a selection of candidate functions to describe the probability of detection and estimates the parameters using maximum likelihood estimation. The probability of detecting an object may not only depend on how far it is from the obsever but also on other factors such as weather conditions, ground cover, cluster size etc. The \pkg{Distance} package also allows the incorporation of such covariates into the detection function allowing the detection function parameters to vary based on these covariates.

Other important assumptions in conventional distance sampling are as follows: all objects exactly on the transect line or point are detected, objects are stationary, measurements are exact. Field methods must therefore be chosen carefully to satisfy these assumptions within the species and habitat constraints [@buckland2015distance].

The rest of the paper has the following structure: we describe data formatting for \pkg{Distance}; candidate detection function models are described in terms of formulation and fitted examples in \proglang{R}. We then show how to perform model checking, goodness of fit and model selection. We go on to show how to estimate abundance, including stratified estimates of abundance. The final two sections of the article look at extensions (both in terms of methodology and software) and put the package in a broader context amongst other software packages used for estimating the abundance of biological populations. (LT: Do we mean software packages in general (e.g., Mark) or R packages?  Do we actually do this -- i.e., talk about how Distance fits in among all this, or is this now an appendix, and does it cover all methods or just distance-based methods.)

>>>>

Until now those wishing to use our \proglang{R} packages for straight forward distance sampling analyses would have had to negotiate the complex package \pkg{mrds} [@mrds-pkg] designed for mark-recapture distance sampling. In addition, \pkg{mrds} requires a complex data structure before analyses can be completed. \pkg{Distance} is a wrapper package around \pkg{mrds} making it easier to get started with basic distance sampling analyses in \proglang{R}. Like the histogram function, the most basic detection function estimation only requires a numeric vector of distances. Here we demonstrate how to use \pkg{Distance} to fit detection functions, perform model checking and selection, and estimate abundance.

@@

Given that we can assume these availability functions there are a number of ways we can use either the estimated detection function or the composite function we fit to our data to estimate abundance. One simple idea begins with finding the area under the composite function out to a truncation distance $w$. If we then find the area under the availability function out to distance $w$, then dividing the former area by the latter gives us an estimate of the proportion of objects seen within distance $w$ of the transect. We can then scale up the number of objects we saw, first to give us an estimate of how many objects were within distance $w$ of the transects and then again by the proportion of the study area we have sampled to give an estimate of abundance for the entire study area.

```{r detectdists, fig.height=5.5, fig.width=8, echo=FALSE, message=FALSE, fig.cap="Top panels show an example detection function (left), the expected availability of objects (middle) and the resulting distribution of detections (right) for line transects. Bottom panels show an example detection function (left), the expected availability of objects (middle) and the resulting distribution of detections (right) for point transects. The distributions in the right hand plots are obtained by multiplying the detection function by the availability function and are referred to in the text as the `composite function'. \\label{fig:detectiondist}"}

@@

Figure \ref{fig:pointslines} shows simulated sampling of a population of 500 objects using line and point transects and their corresponding histograms of observed detection distances. Note that for the purposes of distance sampling an object may either refer to an individual in a population or a cluster of individuals.

@@

\pkg{Distance} provides a selection of candidate functions to describe the probability of detection and estimates the associated parameters using maximum likelihood estimation. The probability of detecting an object may not only depend on how far it is from the obsever but also on other factors such as weather conditions, ground cover, cluster size etc. The \pkg{Distance} package also allows the incorporation of such covariates into the detection function allowing the detection function scale parameter to vary based on these covariates.

In addition to randomly placed samplers distance sampling relies on three other main assumptions. Firstly, all objects directly on the transect line or point (i.e. those at zero distance) are detected (see "Extensions" for methods to deal with the situation when this is not possible). Secondly, objects are stationary or detected prior to any movement. Thirdly, distance to the object must be measured accurately, or the observation allocated to the correct distance bin for grouped data. Depending on your survey species some of these assumptions may be more difficult to meet than others. Further information on field methods to help meet these assumptions can be found in [@buckland2015distance].

The rest of the paper has the following structure: we describe data formatting for \pkg{Distance}; candidate detection function models are described in terms of formulation and fitted examples in \proglang{R}. We then show how to perform model checking, goodness of fit and model selection. We go on to show how to estimate abundance, including stratified estimates of abundance. The final two sections of the article look at extensions (both in terms of methodology and software) and put the package in a broader context amongst other \proglang{R} packages used for estimating the abundance of biological populations. <!-- (LT: Do we mean software packages in general (e.g., Mark) or R packages?  Do we actually do this -- i.e., talk about how Distance fits in among all this, or is this now an appendix, and does it cover all methods or just distance-based methods.) LM: changed this to R packages) -->
